---
title: "spatialwarnings - User Guide"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: false 
    code_folding: show
---

The best way to get an introduction to the package 'spatialwarnings'
is through our freely-available
[paper](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.13058),
which describes the scientific background behind early-warning signals, and how 
they can be computed with the package. The purpose of this document is to 
provide a more technical introduction, and answer more advanced 
questions/issues. Please note that if you make a significant use of our package 
in your work, we would appreciate that you cite the above paper. 

This document assumes some basic familiarity with R code. All the required 
files and source code needed to build this document are available 
[there](https://github.com/spatial-ews/spatialwarnings-faq).

```{r setup, include = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  fig.width = 8,
  fig.height = 6,
  out.width = "700px"
)

# Install some packages which are needed by this vignette
options(repos = "https://cloud.r-project.org")

pkgs <- c('tiff', 'png', 'gridExtra', "acss", 'raster',
          "ggplot2", "plyr", "tidyr", 'rgdal')
# EBImage can be installed from Bioconductor repositories:
# https://www.bioconductor.org/packages/release/bioc/html/EBImage.html
if ( ! require(EBImage) ) {
  install.packages("BiocManager")
  BiocManager::install("EBImage", suppressUpdates = TRUE)
  library(EBImage)
}
for (p in pkgs) { 
  stopifnot(require(p, char = TRUE))
}

```





# A 5-minutes introduction to `spatialwarnings`

As ecological systems degrade along gradients, several "Early-Warning signals" 
have been suggested in the literature to monitor such degradation. In practice, 
they are indicators that measure a property of an ecological system, whose trend is expected to change before an abrupt shift in ecosystem condition along a gradient (or in time). For example, the spatial variability (e.g. variation in vegetation cover) of a given ecosystem is expected to rise before some regime shifts. 

A variety of such spatial indicators have been suggested in the 
literature: the package `spatialwarnings` focuses on computing spatial 
indicators, i.e. those that are based on measures of spatial characteristics of 
an ecosystem. As a result the base type of data used by `spatialwarnings` are 
matrices (2D arrays of numbers), that typically come from raster data (e.g. 
aerial images). For example, using the package, we can compute the spatial 
variance of the following matrix: 

```{r autocorr_example, fig.width = 6, fig.height = 6}

# Load the package and enable parallel processing
# library(spatialwarnings) 
plan(multicore)

data(serengeti) # get more information on this dataset by using ?serengeti

# Extract and display matrix from one of the example datasets
example_matrix <- serengeti[[1]]
display_matrix(example_matrix)

# This function returns the coarse-grained spatial variance when fed a matrix.
# See ?raw_cg_variance for more details about its computation.
raw_cg_variance(example_matrix)

```

When working with indicators, we are often interested in their change along a 
gradient rather than in a single value. A typical use-case is having several 
aerial images (matrices), each corresponding to a given location on a gradient. 
`spatialwarnings` provides all of what is needed for such case. By storing 
matrices in a list, indicators can be computed on all of them at once: 

```{r compute_set, fig.width = 6, fig.height = 6}
trend <- compute_indicator(serengeti, raw_cg_variance)
trend
```

The package has computed the indicator value for all matrices and displayed a 
textual summary. We can now plot the trend along the gradient by calling the 
generic function `plot()` on the returned object. Here the example dataset is 
along a gradient of increasing rainfall, so we will use the rainfall values as our x-axis in a trend plot: 

```{r plot_trend}
plot(trend, along = serengeti.rain) 
```

Because the tree cover changes along the gradient, this may affect the indicator trends, which may then not reflect the proximity to the shift, but rather changes in cover. Using `spatialwarnings`, we whether the indicator values deviate from their null expectation (i.e. their expected value given the vegation cover). This is done by randomizing the spatial structure of the matrix, and recomputing indicator values on such matrix: 

```{r compute_null}
variance_test <- indictest(trend)
variance_test
```

Here all observed values of spatial variance are above their null distribution 
(P-value indistiguishable from zero). Again, we can plot trends and investigate 
how these null values compare to the indicator trends: 

```{r plot_null}
plot(variance_test)
```

The grey ribbon indicates the 0.05/0.95 quantiles of the null distribution (it 
shows up as a thin line on the graph). 

`spatialwarnings` allows computing indicators following such three-step 
workflow. For convenience, some indicators are also available 'umbrella' 
functions. For example, the spatial variance above is part of the so-called 
"generic indicators", which also comprise spatial autocorrelation and skewness. 
We can compute and display all of them at once using the `generic_sews` 
function: 

```{r allatonce}

generic_trend <- generic_sews(serengeti)

# Here we adjust and annotate the plot using ggplot2 directives 
plot(generic_trend, along = serengeti.rain) + 
  geom_vline(xintercept = 733, color = "red", linetype = "dashed") +
  labs(x = "Annual rainfall",
       y = "Mean cover/indicator value",
       title = paste("Early warning signals of a shift in tree cover in", 
                     "Serengeti, Tanzania (Eby et al. 2016)"), 
       subtitle = paste("Grey ribbons indicate the 5-95% quantiles of the null", 
                        "distribution"))

```





# What indicators are available in `spatialwarnings`? 

`spatialwarnings` implements the following indicators found in the literature: 

  - lag-1 autocorrelation (Moran's I, Dakos et al. 2010)
  - Variance (Guttal et al. 2008)
  - Skewness (Guttal et al. 2008)
  - SDR Ratio (Kéfi et al. 2014)
  - Patch-size distribution shape (Kéfi et al. 2011)
  - Power-law range (Berdugo et al. 2017)
  - Flowlength (Rodriguez et al. 2017)
  - Kolmogorov complexity (Dakos et al. 2017)
<!--   - Variogram-based indicators (Nijp et al. 2019) -->

Do not hesitate to get in touch through [Github](https://github.com/spatial-ews/spatialwarnings) if you want to see this list grow ! 





# What do the reported P-values mean? 

The choice made in `spatialwarnings` is to report P-values that are low when the indicator value is above the null distribution. This is motivated by the
fact that most indicators tend to have increasing values near a possible
transition. For example, a P-value of 0.01 means that the observed indicator
value is above 99% of the values observed in the null distribution. 

However, using this convention means that indicators which are expected to show 
significant deviation when below their corresponding null distribution will show apparently non-significant P-values (i.e. close to one). 





# spatialwarnings is slow ! 

While a best effort is made to make `spatialwarnings` as fast as possible, it 
relies heavily on permutating matrices and recomputing indicators, which can 
take a while to compute on personal computers. You can enable parallel processing to alleviate this. In that case, when computing indicators on lists
of matrices, `spatialwarnings` will use one core per matrix. Parallel
processing can be enabled using the function `plan()` (provided by the 
[future](https://cran.r-project.org/package=future) framework): 

```{r parallel}
plan(multisession)
```

Note that the `future` package provides many other ways to parallelize 
computations (e.g. across multiple computers in a cluster). For more 
information, please refer to the [official manual](https://cran.r-project.org/web/packages/future/vignettes/future-1-overview.html).





# How can I deal with `NA`s in the data?

Put simply, there is no single way to deal with missing values, this is why it's up to the user of the package to deal with those before computing indicators. It depends on the meaning of those `NA`s.

If the matrix is has continuous values, then `NA`s generally represent
missing measurements. In this case, interpolation may be an option, but
computing indicators on interpolated data may be lead to wrong conclusions
(because it increases autocorrelation in the data, which is by itself
an indicator !), see e.g.
[this blog post](https://www.fromthebottomoftheheap.net/2013/07/16/flickering-diatoms/).

If the matrix has discrete (TRUE/FALSE) values, then `NA`s may represent either 
(i) missing measurements, and in this case you are in the above case or (ii) a 
pixel that has been observed, but where the value represents the non-focal state (i.e. what should be in fact, a FALSE value). In this case, it should be safe to replace the `NA`s in the matrix by FALSE values.

For example, a matrix could contain `TRUE` values where the pixel contains a
given cover of forest and `NA` if it contains any other thing
(buildings, grassland, etc.). If you are interested in the spatial dynamics
of forest patches, then it should be safe to replace those `NA`s by FALSE
values and compute indicators.





# How can I add another indicator?

'spatialwarnings' provides the function `create_indicator`, which allows the 
user to define new indicators, and use them as if they were `*_sews` functions 
(and thus benefit from significance testing, plotting functions, etc.).

The process is in two steps, we first define in R a function that computes the
metric. This function should take a matrix as input, and return a single,
numerical value. Here, we define a function that returns the maximum patch 
size in a matrix. 


```{r custom_indic}
# This function takes a matrix and a returns a single value.
get_maxpatchsize <- function(mat) {
  c(maxpatchsize = max(patchsizes(mat)))
}
```

We then create an "indicator function" using the above function. This
indicator function can then be used as any other function from the `*_sews`
family. You just define how the metric is derived from data, 'spatialwarnings'
handles all the significance-testing and plotting automatically.

```{r custom_indic_func}
# Create the indicator function
maxpatchsize_sews <- create_indicator(get_maxpatchsize)

# Apply this function on the serengeti dataset. We reduce the number of
# permutations when testing significance to speed up computations. 
maxpatch_indic <- maxpatchsize_sews(serengeti)
maxpatch_indic <- indictest(maxpatch_indic, nulln = 99)

# Display the trends
plot(maxpatch_indic, along = serengeti.rain)

# Display text-based summary
summary(maxpatch_indic)

```

You can also do this on the fly with the function `compute_indicator()`, 
which produces the same output without creating an intermediary function: 

```{r compute_indic}

maxpatch_indic <- compute_indicator(serengeti, get_maxpatchsize)
plot(maxpatch_indic, along = serengeti.rain)

```

If you want to distribute a new indicator with the package, do not
hesitate to open a new topic on our
[Github](https://github.com/spatial-ews/spatialwarnings/issues) tracker!





# Can I use `Raster*` objects with 'spatialwarnings'?

`Raster*` objects are used by the package `raster` to encapsulate raster data, 
for example data read from Geotiff images. These objects can contain a single 
band (`RasterLayer` objects, with one value per pixel) or multiple bands 
(`RasterStack`/`RasterBrick` objects, with several values per pixel). Each band 
of a raster object is, by definition, a matrix of values.

Base `spatialwarnings` cannot use `RasterLayer` objects directly, but a 
companion package `spatialwarningsGis` (available [there](https://github.com/spatial-ews/spatialwarningsGis)) provides extensions to work with these 
objects. You can install the package using devtools: 

```{r install_GIS_exts}
devtools::install_github('spatial-ews/spatialwarningsGis')
library(spatialwarningsGis)
```

You can now use raster objects transparently with `spatialwarnings`, they will 
be converted to matrices internally and without extra work for the user. A workflow with `RasterLayer` object could then look like the following: 

```{r raster_objs}

# It makes no sense computing indicators on those images, they just are
# example geotiffs
example_files <- 
  paste0("https://download.osgeo.org/geotiff/samples/spot/chicago/", 
         c("SP27GTIF.TIF","UTM2GTIF.TIF"))

# Read the list of files using functions from the package 'raster'
list_of_raster_objects <- lapply(example_files, function(f) {
  tmpfile <- tempfile()
  download.file(f, destfile = tmpfile)
  raster(tmpfile)
})

# ...and use them in spatialwarnings
generic_sews(list_of_raster_objects)

```

Right now some metadata of the raster objects is lost in the process
(extent, resolution, etc.). This information could be used by 'spatialwarnings'
in a future release to adjust output (for example, adjust distance units when
displaying the r-spectrum).

Please note that `spatialwarnings` will not use the multi-band objects 
`RasterStack`/`RasterBrick` as the implemented indicators only work with 
univariate values. It is unknown which band should be picked in multi-band 
objects to compute the indicators. You need to extract `RasterLayer`s from these objects first, then use `spatialwarnings`

Additional support for other GIS formats representing raster data will be added
to `spatialwarningsGis`. 





# How can I identify the periodicity in data?

EWS available in `spatialwarnings` assume that stress does not affect the 
periodicity of the spatial structure, as changes in periodicity may mask or 
alter trends in indicator values. Before computing the EWS, it is therefore 
needed to check for any possible periodicity in the input data. In this section, we contrast two images to show how the r-spectrum can be used to
highlight periodicity (or its absence).

```{r computation}

# Read images
image_periodic <- readPNG('./patterned_bush_niger.png')
image_nonper   <- readPNG('./non_patterned_spain.png')

# The images have three bands (R/G/B). We transform them into a black
# and white image using principal component analysis (PCA)
summarise_pca <- function(arr) {
  values <- matrix(as.vector(arr), ncol = 3)
  pca <- prcomp(values)
  values <- pca[["x"]][ ,1]
  matrix(values, ncol = ncol(arr), nrow = nrow(arr))
}

# Transform into one-band (black and white image) and compute r-spectrum.
image_periodic_pca <- summarise_pca(image_periodic)
rspec_periodic <- rspectrum(image_periodic_pca)

image_nonper_pca <- - summarise_pca(image_nonper)
rspec_nonper <- rspectrum(image_nonper_pca)

```

```{r plot_functions, echo = FALSE, print = FALSE}
# Transform a matrix into ggplot-compatible data.frame
mat2longdf <- function(mat) {
  data.frame(expand.grid(y = seq.int(ncol(mat)),
                         x = seq.int(nrow(mat))),
             value = as.vector(mat))
}

# Counter to add letters to figures
counter <- 0
lettercount <- function() letters[counter <<- counter + 1]

# Make series of breaks/labels for log scale
brks <- as.vector(outer(seq.int(9), 10^seq(-10, 10)))
brks.major <- brks[grepl("^1",  brks, perl = TRUE)]
brks.minor <- brks[!grepl("^1", brks, perl = TRUE)]
lbls <- ifelse(grepl('^(1|2|5)', brks.major, perl = TRUE), brks.major, "")
scale_x <- scale_x_continuous(trans = "log10",
                              breaks = brks.major[brks.major>0],
                              minor_breaks = brks.minor[brks.minor>0],
                              labels = lbls[brks.major>0])
scale_y <- scale_y_continuous(trans = "log10",
                              breaks = brks.major,
                              minor_breaks = brks.minor,
                              labels = lbls)
```

The periodic image below (tiger bush in Niger [^1]) displays a typical banded
regular pattern (a). Its scale (in number of pixels) is reflected by a hump in
the r-spectrum around a size of ~ 15px (b). This corresponds to the width of one vegetation band in the image (~30m in reality).

[^1]: https://www.google.com/maps/place/12%C2%B030'47.3%22N+3%C2%B000'49.9%22E/@12.5131562,3.0116833,905m/data=!3m2!1e3!4b1!4m5!3m4!1s0x0:0x0!8m2!3d12.513151!4d3.013872?hl=en]

```{r display_periodic, fig.width = 12, fig.height = 6}

img <- ggplot(mat2longdf(image_periodic_pca)) +
        geom_raster(aes(x = x, y = y, fill = value)) +
        scale_fill_gradient(low = "black", high = "white") +
        scale_y_reverse() +
        theme_minimal() +
        theme(legend.position = "none") +
        labs(x = "Pixel coordinate",
            y = "Pixel coordinate",
            title = paste0(lettercount(), ") Image of periodic vegetation"))

rsp <- ggplot(rspec_periodic) +
         geom_point(aes(x = dist, y = rspec), pch = 20) +
         scale_x + scale_y +
         theme_minimal() +
         labs(x = "Distance (number of pixels)",
             y = "r-spectrum value (power)",
             title = paste0(lettercount(), ") r-spectrum of periodic image"))

gridExtra::grid.arrange(img, rsp, ncol = 2)

```

This image displays strong periodicity and the indicators in `spatialwarnings`
should be used with care to draw conclusions from it.

In contrast, an aperiodic image displays no such hump in its r-spectrum. The 
image below (c, arid grassland in Spain) is aperiodic, as shown by the 
decreasing r-spectrum (d). Indicators available in `spatialwarnings` may (in 
principle) be used on this image.

```{r display_nonperiodic, fig.width = 12, fig.height = 6}

img <- ggplot(mat2longdf(image_nonper_pca)) +
        geom_raster(aes(x = x, y = y, fill = value)) +
        scale_fill_gradient(low = "black", high = "white") +
        scale_y_reverse() +
        theme_minimal() +
        theme(legend.position = "none") +
        labs(x = "Pixel coordinate",
            y = "Pixel coordinate",
            title = paste0(lettercount(), ") Image of non-periodic vegetation"))

rsp <- ggplot(rspec_nonper) +
         geom_point(aes(x = dist, y = rspec), pch = 20) +
         scale_x + scale_y +
         theme_minimal() +
         labs(x = "Distance (number of pixels)",
             y = "r-spectrum value (power)",
             title = paste0(lettercount(), ") r-spectrum of non-periodic image"))

gridExtra::grid.arrange(img, rsp, ncol = 2)

```





# Converting multi-band images

Raster images from remote sensing often come with multiple bands, because of
the way images are stored digitally (e.g. red/blue/green bands), but also
because multiple sensors can be used to capture a single image (e.g. a
combination of visible/near-infrared sensors). These images are often
represented as arrays in R (matrices that are extended with a third dimension,
"depth").

As `spatialwarnings` computes indicators on `matrix` objects, these arrays need
to be first transformed into matrices (effectively arrays with no depth). This
requires collapsing the multi-dimensional data of each pixel to a single value.
In addition, some indicators (e.g. patch-based indicators) rely on classified,
`TRUE/FALSE` data that defines which pixels belong to a patch and which do not.
The classification algorithm employed is likely to be very system-dependent,
but we provide a possible example below.

In what follows, we use a small example to show how to carry these two
processes and apply the `spatialwarnings` package on images. Using an aerial
image stored as an R/G/B PNG image, we classify each pixel as 
vegetated/non-vegetated or summarize the multi-channel pixel values to a 
single value. We then apply the indicators on the resulting matrix.

Note that in many cases, other methods can be used to derive univariate indices 
from remote sensing data that may be more appropriate. For example, a typical 
one-dimensional index for vegetation data is NDVI (Normalized Difference 
Vegetation Index), which measures the extent of green vegetation in a given 
pixel and is obtained as the normalized difference between the red a 
near-infrared bands of a multi-band image.

## Classifying an RGB image to compute patch-based indicators

We first load the required packages and read the image in R.

```{r pkgs}

# Read image
datadir <- "../figs/images/"
img1 <- readPNG('./crau_quercus_encroachment.png')

# Display image
grid::grid.raster(img1)

```

The vegetation in this image displays some patchiness, that can be characterized by `spatialwarnings`. `img1` is a raster image (represented as an
array in R) with three (red/blue/green) channels that need to be be classified
into `TRUE` (pixel is in a vegetation patch) or `FALSE` values in order to
compute patch-based compute indicators.

```{r densities, fig.width = 16, fig.height = 6}
# Convert the image to a data.frame
img1_tab <- data.frame(expand.grid(x = seq.int(nrow(img1)),
                                   y = seq.int(ncol(img1))),
           as.data.frame(matrix(img1, ncol = 3)))
names(img1_tab) <- c('x', 'y', 'red', 'green', 'blue')

# A very bright object is present in the image, that distorts the results
# further down and needs to be removed. We set its pixel values to NAs.
luminance <- with(img1_tab, sqrt( 0.299*red^2 + 0.587*green^2 + 0.114*blue^2 ))
img1_tab[ ,c('red', 'green', 'blue')] <-
  img1_tab[ ,c('red', 'green', 'blue')] * ifelse(luminance > .5, NA, 1)

```

We can display the distribution of each channel values:

```{r channel_values_distribution}
# fix level order so colors match in graph
df <- gather(img1_tab, channel, value, red, green, blue)
df[ ,'channel'] <- factor(df[ ,'channel'], levels = c("red", "green", "blue"),
                          ordered = TRUE)
ggplot( df ) +
  geom_density(aes(x = value, color = channel)) +
  theme_minimal() +
  labs(caption = "Distribution of channel values") +
  scale_color_manual(values = c('red', 'green', 'blue', 'black'))
```

Each individual channel can be also represented as a monochrome image:

```{r each_channel_independently}
ggplot(gather(img1_tab, channel, value, red, green, blue)) +
  geom_raster(aes(x = x, y = y, fill = value)) +
  facet_grid( ~ channel, labeller = label_both) +
  coord_fixed() +
  theme_minimal() +
  scale_fill_gradient(low = "#000000", high = "#FFFFFF") +
  labs(caption = "Monochrome representation of the three channels of the RGB image")
```

We can classify this image using an unsupervised k-means classification
algorithm on the pixel data. Many other classification algorithms exist and may
show better accuracy, but k-means is simple, generic and fast.

```{r kmeans}

km <- kmeans(na.omit(img1_tab[ ,c('red', 'green', 'blue')]),
             centers = 2)
img1_tab[ ,'clust'] <- NA
img1_tab[!is.na(img1_tab[ ,'red']), 'clust'] <- km[['cluster']]

# The number identifying each cluster is random in k-means: we make sure that
# cluster 2 always has greener values, and thus always corresponds to potential
# vegetation patches.
img1_tab[ ,'clust'] <- with(img1_tab, as.integer(reorder(as.factor(clust), -green)))

ggplot(img1_tab) +
  geom_raster(aes(x = x, y = y,
                  fill = as.factor(clust))) +
  coord_fixed() +
  theme_minimal() +
  scale_fill_manual(values = c('#F4EAA4', '#0A8E0B'))

```

Cluster 2 roughly identifies vegetation patches, but many single-pixel values
are misclassified by the k-means algorithm. However, they most likely pertain to the nearest patch: we apply a smoothing filter to get rid of them.

```{r filter}

clust_filt <- with(img1_tab,
                   gblur(matrix(!is.na(clust) & clust == 2,
                                nrow = max(x), ncol = max(y)),
                                sigma = 1.5)) > .2

img1_tab[ ,'clust_filt'] <- as.vector(clust_filt)

ggplot(img1_tab) +
  geom_raster(aes(x = x, y = y,
                  fill = as.factor(clust_filt))) +
  coord_fixed() +
  theme_minimal() +
  scale_fill_manual(name = "Vegetation",
                    values = c('#F4EAA4', '#0A8E0B'))

```

We can then apply the patch-based indicators on the resulting classified
matrix.

```{r psd_indics}

psd_indic <- patchdistr_sews(clust_filt, fit_lnorm = TRUE)

summary(psd_indic)

plot_distr(psd_indic, best_only = FALSE) +
  labs(title = "Results of patch-size distribution fitting")

```

Note that this brief demonstration uses k-means, but many classification methods exist, either supervised or unsupervised (Liu and Mason, 2016) and may
show better results depending on the focal system.


## Reducing the dimensionality of pixel data

Indicators require each pixel to have a single, uni-dimensional value. One
option is to use the values of each channel independently (e.g. using red values only), but these may not represent very well the biological variable 
of interest in each pixel. 

Similar to classifications, supervised and unsupervised methods exist to
summarize three-dimensional information to a single value. A simple generic
unsupervised method is Principal Component Analysis (see e.g. Liu and Mason,
2016 for more details about its application on raster images), which we
use here as example.

We carry out a PCA on the pixel data:

```{r pca}

# We use the red/green/blue information and combine them using PCA
NApixel <- is.na(img1_tab[ ,'red'])
pca <- prcomp(img1_tab[!NApixel, c('red', 'green', 'blue')])

summary(pca)

# Import data back into data.frame
img1_tab[ ,'pca1'] <- NA
img1_tab[!NApixel,'pca1'] <- predict(pca)[ ,1]

ggplot(img1_tab) +
  geom_raster(aes(x = x, y = y, fill = pca1)) +
  coord_fixed() +
  theme_minimal()

```

Results from PCA show that the first axis explains 98% of the variance in RGB
value, so it summarizes very well the variations in pixel values. We can then
apply indicators on these univariate values (we use here the example of spectral indicators).

Note that because there are some missing values (`r sprintf("%1.2f%%",
round(mean(is.na(img1_tab$pca)), 4)* 100)` of all pixels) in the input image
(the pixels where a very bright object is present), we need to interpolate them
before computing the indicators. Note that interpolation increases the
autocorrelation in a given matrix and indicators should be interpreted with
caution if a large amount of pixels have been interpolated.

```{r indic}

# Convert to matrix format then compute indicators. Because there are NA
# values in the input image, we need to interpolate these values. We fill the
# pixels with NAs with the mean value of their direct neighbors.
pca_mat <- with(img1_tab, matrix(pca1, ncol = max(y), nrow = max(x)))
while ( any( is.na(pca_mat) ) ) {
  for (i in seq.int(nrow(pca_mat))) {
    if ( any(is.na(pca_mat[i, ])) ) {
      for (j in seq.int(ncol(pca_mat))) {

        # Bound check on the matrix
        i2 <- min(i, 1, nrow(pca_mat))
        j2 <- min(j, 1, ncol(pca_mat))
        if ( is.na(pca_mat[i, j]) ) {
          pca_mat[i, j] <- mean(pca_mat[(i2-1):(i2+1),
                                        (j2-1):(j2+1)])
        }
      }
    }
  }
}
```

We can then compute the indicators on the resulting matrix:

```{r indicator_compute}

# Compute spectral indicators
ic <- spectral_sews(pca_mat,
                    sdr_low_range = c(0, .2),
                    sdr_high_range = c(.8, 1))
ic <- indictest(ic, nulln = 499)

# Display textual summary
summary(ic)

```

Note that while PCA may summarize well arbitrary variations in R/G/B pixel
values, this does not mean that the resulting pixel values always capture well
the variations in the system state. Before interpreting indicators, it is
necessary to understand well how the numerical values in the image reflect the
state of the ecological system. The use of indices known to reflect a specific
aspect of the system (e.g. NDVI reflecting the amount of green vegetation cover) may be often more informative than using a generic algorithm to collapse
multiple channels into one.





# Using advanced null models 

To test the significance of indicator values, by default `spatialwarnings` will 
take the original matrix and derive many 'null' matrices from it. A typical
null matrix is one that has the same average value, but a completely random
spatial structure. For example, such a matrix can be obtained by randomizing
all cells in the original matrix. By recomputing the indicators on the null
matrices, we get a distribution of values to which we can compare the observed
indicator value. This way we check if the observed indicator value is consistent with the null hypothesis of no spatial structure, or if there is something significant in the landscape. 

In `spatialwarnings`, the default method used to produce null matrices is to 
reshuffle the cells randomly in the matrix (method `perm` in the arguments of 
`indictest`, see `?indictest`). This will keep the average value of the matrix 
(i.e. the proportion of TRUE cells when a matrix is made of `TRUE`/`FALSE` 
values), but removes any spatial structure. It is the default because it is 
reasonably fast. 

Another method (method `intercept`) will fit an intercept-only generalized 
linear model using `glm()` to the values of the matrix, then use this model to 
predict null values in each cell. Let's consider the following matrix: 

```{r exmat}
indic_example <- generic_sews(serengeti[[13]])
display_matrix(indic_example)
```

We use the 'intercept' method and retrieve a null matrix from the `indictest`
object: 

```{r exmod}

test_example <- indictest(indic_example, nulln = 49, 
                          null_method = "intercept")
display_matrix( test_example$get_nullmat() )

```

This method randomizes the spatial structure, but the cover is preserved. Note 
that this is is very similar to shuffling the position of the cells in the 
matrix. The difference is that the `intercept` method will preserve the cover 
*on average* (for a large number of null matrices), while the `perm` method will preserve the cover exactly. 

When fitting the generalized linear model, `spatialwarnings` tries to make 
a sensible default choice for the `glm()` family. It is crucial that this 
argument is well-chosen, as otherwise the fitted model may be inappropriate. 
By default, `spatialwarnings` uses a `binomial()` model 'family' (`family` 
argument in glm) for logical matrices, and a `gaussian()` family for real-valued matrices. A warning is produced when this automatic choice is made, and 
specifying explicitely the model family will remove the warning: 

```{r model_nowarning}

test_example <- indictest(indic_example, nulln = 49, 
                          null_method = "intercept", 
                          null_control = list(family = binomial()))

```


These two null models assume that all the spatial patterning found in a matrix 
is related to the proximity to a possible regime shift, i.e. *without influence 
of any external driver*. In other words, the null model assumes that the 
probability of finding a given value in the null matrix is constant over the 
whole matrix. This may be an unrealistic assumption for many real-world cases. 

For example, let's consider a case where vegetation cannot grow in some part of 
a landscape because of higher elevation, or the presence of a water body. In 
this case, spatial autocorrelation increases because vegetation is clustered 
into the remaining areas, where plants can grow. Such changes in correlation 
will change indicator values in ways that are unrelated to regime shifts, and 
lead to mis-interpretation. It is thus critical for correct interpretation of 
patterns to make sure no unaccounted-for variable drives changes in indicator
values. 

There is no "one size fits all" method to take care of such confounding 
environmental driver. Ideally, one would estimate the probability of observing a given value in a cell (e.g. vegetation cover), given the environmental
drivers in that cell, but without the effect of the biological processes
producing the regime shifts. This requires having detailed spatial data, i.e.
not only the matrix with observed values, but also covariates driving these
observed values (e.g. rainfall maps). This is not impossible to do but it is
very much case-specific. 

A possible null model of intermediate complexity, that does not need external 
data, is assuming that any large-scale changes in values are driven by 
environmental factors, while small-scale changes are related to biological 
processes (and thus reflect better the changes in indicator values related to 
regime shifts). To do so, we can estimate changes in matrix values over large 
scales, and use those estimated probabilities to produce the null matrices. 

Let's consider the previous matrix (describing the vegetation cover in
serengeti): 

```{r serengeti_example}
display_matrix(indic_example)
```

It looks like overall, there is more vegetation in the bottom part of the matrix than in its upper part. In other words, a cell has a higher probability
of being covered by forest when it is on the bottom side of the matrix. Such
large-scale pattern could be related to a change in rainfall or altitude,
rather than proximity to a regime shift. It would be good to take it into
account to make sure it does not drive the trends in autocorrelation. 

We can do so by fitting a smoothed version of the landscape over large scales, 
using `gam()`. This model will allow the average value of a cell to change 
depending on the position of the cell in the matrix. Here we use this null 
model and draw a random, 'null' matrix: 

```{r modsmooth}

test_example <- indictest(indic_example, nulln = 99, 
                          null_method = "smooth")
display_matrix( test_example$get_nullmat() )

```

As expected, the model picks up the fact that the bottom side of the landscape 
tends to have more forest: in those areas, the null matrix has a higher 
probability of being covered. 

Let's compare the three null models using the 'generic indicators': 

```{r null_comparision}

indics <- generic_sews(serengeti)

control <- list(familiy = binomial())

# Compute and export the values 
all_methods <- ldply(c("perm", "intercept", "smooth"), function(method) { 
  test <- indictest(indics, nulln = 99, 
                    null_method = method, 
                    null_control = control)
  data.frame(method = method, as.data.frame(test))
})

# Display the results
all_methods[ ,"x"] <- serengeti.rain[all_methods[ ,"matrixn"]]
ggplot(all_methods, aes(x = x)) + 
  geom_line(aes(y = null_mean, group = method), alpha = .2) + 
  geom_ribbon(aes(ymin = null_qinf, ymax = null_qsup, fill = method), 
              alpha = .8) + 
  geom_point(aes(y = value), size = 1) + 
  geom_line(aes(y = value)) + 
  facet_wrap( ~ indic, scales = "free_y") + 
  theme_minimal() + 
  labs(x = "Rainfall (mm)", 
       y = "Indicator value")

```

In the above figure, the observed indicator values are represented with the 
black line. The null values are represented by the ribbons, with the color 
indicating the null method used. The 'intercept' and 'perm' values give very 
similar results. The 'smooth' method produces values that are closer to the 
observed indicator value, i.e. it is more conservative than the shuffling or 
intercept-only model. This is expected as the smooth method produces null 
matrices that are closer to the original matrix. 

When running analyses, it is probably worth trying these different sets of null 
models to make up your mind about how strongly values deviate from random 
expectations. 







# References

Dakos, V., and F. Soler-Toscano. 2017. Measuring complexity to infer changes in
the dynamics of ecological systems under stress. Ecological Complexity
32:144–155.

Liu and Mason, 2016. Image Processing and GIS for Remote Sensing: Techniques and Applications. John Wiley \& Sons, Ltd.
